---
title: "VCM: COVID Rates Analysis"
author: "Benjamin Garcia"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

**Getting started**

```{r, message = FALSE, warning = FALSE}
# Clear workspace
rm(list = ls())

# Load required libraries
library(dplyr)       # For data manipulation
library(ggplot2)     # For plotting
library(tidyverse)   # pivot_longer
#library(tidyverse)

```

**COVID rates data**

```{r, message = FALSE, warning = FALSE}
# Load COVID data
# source: https://data.cdc.gov/Public-Health-Surveillance/Monthly-Rates-of-Laboratory-Confirmed-COVID-19-Hos/cf5u-bm9w/about_data
dataCOVID <- read.csv("Monthly_Rates_of_Laboratory-Confirmed_COVID-19_Hospitalizations_from_the_COVID-NET_Surveillance_System_20250303.csv")

```

```{r}
head(dataCOVID)

# Add 'Month' variable (Month = 1 indicates March 2020, 2 indicates April 2020, ...)
Year <- as.numeric(substr(dataCOVID$X_YearMonth, 1, 4))
MonthNum <- as.numeric(substr(dataCOVID$X_YearMonth, 5, 6))
base_year  <- 2020
base_month <- 3
dataCOVID$Month <- 12 * (Year - base_year) + (MonthNum - base_month) + 1

```

```{r}
summary(as.factor(dataCOVID$Month))
```


```{r}
# Let's check how the following variables look like: AgeCategory_Legend, Sex_Label, Race_Label
unique(dataCOVID$AgeCategory_Legend)
# Category 1: "All" (1 class)
# Category 2: "0-17 years (Children)", "≥18 years (Adults)" (2 classes)
# Category 3: "0-<1 year", "1-4 years", "5-17 years", "18-49 years", "50-64 years", "≥65 years" (6 classes)
# "0-<6 months", "6mo-<12 months", "1-<2 years", "2-4 years", "5-11 years", "12-17 years", "18-29 years", "30-39 years", "40-49 years", "50-64 years", "65-74 years", "75-84 years", "≥85 years" (13 classes)
# there are two remaining classes, which may be used in Category 1 or 2: "0-4 years", "≥75 years"

```

```{r}
summary(as.factor(dataCOVID$State))
```



```{r}
unique(dataCOVID$Sex_Label)
# Category 1: "All"
# Category 2: "Female", "Male"

```

```{r}

unique(dataCOVID$Race_Label)
# Category 1: "All"
# Category 2: "Hispanic", "A/PI, non-Hispanic", "AI/AN, non-Hispanic", "Black, non-Hispanic", "White, non-Hispanic"

```

```{r}

# For now let's aggregate the data over all age, sex, race
library(dplyr)
data_aggregated <- dataCOVID %>%
  filter(AgeCategory_Legend == "All",
         Sex_Label == "All",
         Race_Label == "All") %>%
  group_by(Month) %>%
  summarize(MonthlyCOVIDRate = mean(MonthlyRate, na.rm = TRUE))

```


```{r}

# Consider harmonic terms for seasonal effects
# updated seasonal effects (removed second pi)
data_aggregated <- data_aggregated %>%
  mutate(
    s1 = cos(2 * pi * Month / 12),
    s2 = sin(2 * pi * Month / 12),
    s3 = cos(2 * pi * Month / 6),
    s4 = sin(2 * pi * Month / 6)
  )

```


```{r}
# s1 s2 annual cycle
#s3 S4 semiannual  
# Visualize the harmonic terms
harmonics <- data_aggregated %>%
  pivot_longer(cols = c(s1, s2, s3, s4), names_to = "term", values_to = "value")
ggplot(harmonics, aes(x = Month, y = value, color = term)) +
  geom_line() +
  scale_color_manual(values = c("s1" = "black", "s2" = "red", "s3" = "green", "s4" = "blue")) +
  labs(x = "Month", y = "Value", title = "Harmonic Terms over Month") +
  theme_minimal()

```


```{r}
#danny <- data.frame(
#  day = as.Date("2020-03-01") + c(0,30,60),
#  value = runif(3) + seq(-140, -138)^2 / 10000
#)
#runif(6)
# Most basic bubble plot
#p <- ggplot(danny, aes(x=day, y=value)) +
#  geom_line() + 
#  xlab("")
#p

 start_date <- as.Date("2020-03-01")
    end_date <- as.Date("2025-02-01")
    monthly_dates <- seq(start_date, end_date, by = "1 month")
    print(monthly_dates)

```



```{r}
data_aggregated$date <- monthly_dates
```


```{r}

# Visualize the COVID rates
ggplot(data_aggregated, aes(x = date, y = MonthlyCOVIDRate)) +
  geom_line() +
  labs(x = 'Date', y = "COVID-NET Hospitalization Rate") +
  ggtitle("Aggregate Average COVID-NET Monthly Hospitalization Rate")+
  scale_x_date(date_breaks = '6 months', date_labels = "%m-%Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(face="bold", size = 11)) +
  theme(axis.text.y = element_text(face = 'bold', size = 11))
```


```{r}
ggplot(data_aggregated, aes(x = Month, y = MonthlyCOVIDRate)) +
  geom_line() +
  labs(x = 'Date', y = "COVID-NET Hospitalization Rate") +
  ggtitle("Aggregate Average COVID-NET Monthly Hospitalization Rate")+
  #scale_x_date(date_breaks = '6 months', date_labels = "%m-%Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(face="bold", size = 11)) +
  theme(axis.text.y = element_text(face = 'bold', size = 11))
```


```{r}
datcov3 <- dataCOVID[dataCOVID$AgeCategory_Legend == 'All',]
datcov3 <- datcov3[datcov3$Sex_Label == 'All',]
datcov3 <- datcov3[datcov3$Race_Label == 'All',]
mean(datcov3[datcov3$Month == 60,7], na.rm = T)

```


**Flu rates data**

```{r, message = FALSE, warning = FALSE}
dataFlu <- read.csv("FluSurveillance_Custom_Download_Data.csv", skip = 2)

dataFlu <- dataFlu[1:(nrow(dataFlu) - 29),] # exclude the description attached at the bottom


```



```{r}
head(dataFlu)

# Convert week to month
dataFlu <- dataFlu %>%
  mutate(
    month = case_when(
      WEEK >= 40 & WEEK <= 43 ~ 10,
      WEEK >= 44 & WEEK <= 47 ~ 11,
      WEEK >= 48 & WEEK <= 53 ~ 12,
      WEEK >= 1  & WEEK <= 4  ~ 1,
      WEEK >= 5  & WEEK <= 8  ~ 2,
      WEEK >= 9  & WEEK <= 12 ~ 3,
      WEEK >= 13 & WEEK <= 17 ~ 4,
      WEEK >= 18 & WEEK <= 21 ~ 5,
      WEEK >= 22 & WEEK <= 26 ~ 6,
      WEEK >= 27 & WEEK <= 30 ~ 7,
      WEEK >= 31 & WEEK <= 35 ~ 8,
      WEEK >= 36 & WEEK <= 39 ~ 9,
      TRUE ~ NA_real_  # fallback if none match
    )
  )


```

```{r}
summary(as.factor(dataFlu$YEAR.1))
dataFlu <- dataFlu[dataFlu$YEAR.1 !='2019',]
#dataFlu <- dataFlu[dataFlu$]

```

```{r}
dataFlu <- dataFlu[(dataFlu$YEAR.1 != '2020') | (dataFlu$month > 2),]

# eliminate jan and feb in 2020
```



```{r}
summary(as.factor(dataFlu$YEAR.1))
summary(as.factor(dataFlu$month))
table(as.factor(dataFlu$YEAR.1), as.factor(dataFlu$month))
```

```{r}
# eliminate march and april of 2025 (had no occurred at the time of data extraction)

dataFlu <- dataFlu[(dataFlu$YEAR.1 != '2025') | (dataFlu$month < 3),]
```


```{r}
table(as.factor(dataFlu$YEAR.1), as.factor(dataFlu$month))
```



```{r}
# Add 'Month' variable (Month = 1 indicates March 2020, 2 indicates April 2020, ...)
Year <- as.numeric(dataFlu$YEAR.1)
MonthNum <- as.numeric(dataFlu$month)
base_year  <- 2020
base_month <- 3
dataFlu$Month <- 12 * (Year - base_year) + (MonthNum - base_month) + 1

```

```{r}
summary(as.factor(dataFlu$Month))
```
```{r}
summary(as.factor(dataFlu$CATCHMENT))
head(dataFlu[dataFlu$Month == 62,])
# month 61 and 62 
```

```{r}
summary(as.factor(dataFlu$WEEK))
```


```{r}

# For now let's aggregate the data over all age, sex, race, virus type
dataFlu$WEEKLY.RATE[dataFlu$WEEKLY.RATE == "null"] <- NA
dataFlu$WEEKLY.RATE <- as.numeric(dataFlu$WEEKLY.RATE)

dataFlu_aggregated <- dataFlu %>%
  filter(AGE.CATEGORY == "Overall",
         SEX.CATEGORY == "Overall",
         RACE.CATEGORY == "Overall",
         VIRUS.TYPE.CATEGORY == "Overall") %>%
  group_by(Month) %>%
  summarize(MonthlyFluRate = sum(WEEKLY.RATE, na.rm = TRUE))

```



```{r}
# imput data

library(imputeTS)

nr1 <- data.frame(Month = c(3,4,5,6,7), MonthlyFluRate = c(NA, NA, NA, NA, NA))
nr2 <- data.frame(Month = c(15, 16, 17, 18, 19), MonthlyFluRate = c(NA, NA, NA, NA, NA))
nr3 <- data.frame(Month = c(29, 30, 31), MonthlyFluRate = c(NA, NA, NA))
nr4 <- data.frame(Month = c(39, 40, 41, 42, 43), MonthlyFluRate = c(NA, NA, NA, NA, NA))
dataFlu_aggregated <- rbind(dataFlu_aggregated[1:2, ], nr1, dataFlu_aggregated[3:9, ], nr2, dataFlu_aggregated[10:18,],
                            nr3, dataFlu_aggregated[19:25,], nr4, dataFlu_aggregated[26:42,])
#imputeTS::na_se
# Decomposes time series, imputes trend/seasonal separately
#fluff <- na_seadec(flu_data$hospitalizations, algorithm = "interpolation")
```

```{r}
#imputeTS::na_se
# Decomposes time series, imputes trend/seasonal separately
dataFlu_aggregated$MonthlyFluRate <- na_seadec(dataFlu_aggregated$MonthlyFluRate, algorithm = "interpolation")
```



```{r}
# Visualize the flu rates
ggplot(dataFlu_aggregated, aes(x = data_aggregated$date , y = MonthlyFluRate)) +
  geom_line() +
  labs(x = "Month", y = "Monthly Flu Rate") +
  scale_x_date(date_breaks = '6 months', date_labels = "%m-%Y")+
  ggtitle("Aggregate Average Flu Monthly Hospitalization Rate")+
  theme_minimal() +
  theme(axis.text.x = element_text(face="bold", size = 11)) +
  theme(axis.text.y = element_text(face = 'bold', size = 11))
```

**rsv rates data**

```{r, message = FALSE, warning = FALSE}
dataRSV <- read.csv("Monthly_Rates_of_Laboratory-Confirmed_RSV_Hospitalizations_from_the_RSV-NET_Surveillance_System_20250401.csv")

```


```{r}
head(dataRSV)

```

```{r}
colnames(dataRSV)[3] <- "Year_Month"
```


```{r}

# Add 'Month' variable (Month = 1 indicates March 2020, 2 indicates April 2020, ...)
Year <- as.numeric(substr(dataRSV$Year_Month, 1, 4))
MonthNum <- as.numeric(substr(dataRSV$Year_Month, 5, 6))
base_year  <- 2020
base_month <- 3
dataRSV$Month <- 12 * (Year - base_year) + (MonthNum - base_month) + 1
```


```{r}
summary(as.factor(dataRSV$Month))
head(dataRSV[dataRSV$Month == -40,])
```

```{r}
dataRSV <- dataRSV[(dataRSV$Year_Month > '202002') & (dataRSV$Year_Month < '202503'),]
```


```{r}
summary(as.factor(dataRSV$Month))
```

```{r}
unique(dataRSV$Age.Category)
summary(as.factor(dataRSV$State))
unique(dataRSV$Sex)
unique(dataRSV$Race)
```
```{r}
library(dplyr)
dataRSV_aggregated <- dataRSV %>%
  filter(Age.Category == "All",
         Sex == "All",
         Race == "All") %>%
  group_by(Month) %>%
  summarize(MonthlyRSVRate = mean(Rate, na.rm = TRUE))
```


```{r}
nr1rsv <- data.frame(Month = c(3,4,5,6,7), MonthlyRSVRate = c(NA, NA, NA, NA, NA))

dataRSV_aggregated <- rbind(dataRSV_aggregated[1:2, ], nr1rsv, dataRSV_aggregated[3:55, ])
```


```{r}
dataRSV_aggregated$MonthlyRSVRate <- na_seadec(dataRSV_aggregated$MonthlyRSVRate, algorithm = "interpolation")
```



```{r}
ggplot(dataRSV_aggregated, aes(x = data_aggregated$date, y = MonthlyRSVRate)) +
  geom_line() +
  labs(x = "Month", y = "Monthly RSV Rate") +
  scale_x_date(date_breaks = '6 months', date_labels = "%m-%Y")+
  ggtitle("Aggregate Average RSV Monthly Hospitalization Rate")+
  theme_minimal() +
  theme(axis.text.x = element_text(face="bold", size = 11)) +
  theme(axis.text.y = element_text(face = 'bold', size = 11))
```



**Combine three data sets**

```{r, message = FALSE, warning = FALSE}
# Select overlapped Month
combineddata <- inner_join(data_aggregated, dataFlu_aggregated, by = "Month")
combined_data <- inner_join(combineddata, dataRSV_aggregated)
## Make the smallest Month -> 1 (not neccessary)
#combined_data <- combined_data %>%
#  mutate(Month = Month - 7)

# Explanatory data analysis:
# check scatter plot and see the relationship between predictor and response over time.
library(ggplot2)
ggplot(combined_data,
       aes(x = MonthlyFluRate, y = MonthlyCOVIDRate, color = Month)) +
  geom_point() +
  theme(axis.text.x = element_text(face="bold", size = 11)) +
  theme(axis.text.y = element_text(face = 'bold', size = 11))
```

```{r}
# in the early months or dark circles we see as there is an increase in covid rate there is only a little increase in flu rate. In later months, as flu rate increases there is not much increase in covid rates. Per 100,000
```


```{r}
library(ggplot2)
ggplot(combined_data,
       aes(x = MonthlyRSVRate, y = MonthlyCOVIDRate, color = Month)) +
  geom_point() +
  theme(axis.text.x = element_text(face="bold", size = 11)) +
  theme(axis.text.y = element_text(face = 'bold', size = 11))
```


```{r}
qnorm(.9) # qnorm is probability convert to z-score
dnorm(0) #is pdf value at that specific points
rnorm(8) # random number generation from mean 0 sd 1
pnorm(1.28) # pnorm is z-score converted to probability to the left
```


**Modeling**

```{r, message = FALSE, warning = FALSE}
# ESTIMATION: VCM

# Kernel function
my.w <- function(X, type = "epanechnikov"){
  if(type == "gaussian"){
    wts <- dnorm(X)
  } else if(type == "epanechnikov"){
    wts <- (3/4) * (1 - X^2) * ifelse(1 < X^2, 0, 1)
  }
  return(wts)
}

# My own VCM function
vcm <- function(y, X, type = "epanechnikov", t, grid, bandwidth){
  num.grid <- length(grid)
  n <- nrow(X); k <- 1 + ncol(X)
  y.vec <- matrix(y, length(y), 1)
  X.mat <- as.matrix(cbind("intercept" = 1, X))
  coef.res <- se.res <- matrix(NA, num.grid, k); mse <- 0
  for(i in 1:num.grid){
    U.star <- as.matrix(diag(t - grid[i], n, n))
    Gamma.mat <- cbind(X.mat, U.star %*% X.mat)
    wts <- my.w((t - grid[i])/bandwidth, type)/bandwidth
    W <- as.matrix(diag(wts, n, n))
    est.beta <- solve(t(Gamma.mat) %*% W %*% Gamma.mat) %*%
      (t(Gamma.mat) %*% W %*% y.vec)
    coef.res[i, ] <- est.beta <- diag(1, k, 2 * k) %*% est.beta
    sigma.hat <- sum((y - X.mat %*% est.beta)^2 * wts)/sum(wts)
    Cov.mat <- solve(t(Gamma.mat) %*% W %*% Gamma.mat) %*%
      (t(Gamma.mat) %*% W %*% W %*% Gamma.mat) %*%
      solve(t(Gamma.mat) %*% W %*% Gamma.mat) * sigma.hat
    se.res[i, ] <- diag(Cov.mat[1:k, 1:k])
    mse <- mse + mean((y - X.mat %*% est.beta)^2)
  }
  return(list(mse = mse/num.grid, coef = coef.res, se = se.res))
}

```


```{r}
combined_data <- combined_data %>% relocate(date, .after = MonthlyRSVRate)

```


```{r}
#shapiro.test(combined_data$MonthlyCOVIDRate)
#plot(density(combined_data$MonthlyCOVIDRate))
```

```{r}
#shapiro.test(combined_data$s1)
#plot(density(combined_data$s1))
```

```{r}
#shapiro.test(combined_data$s3)
#plot(density(combined_data$s3))
```

```{r}
#shapiro.test(combined_data$MonthlyFluRate)
#plot(density(combined_data$MonthlyFluRate))
```


```{r}
#shapiro.test(combined_data$MonthlyRSVRate)
#plot(density(combined_data$MonthlyRSVRate))
```




```{r}
# LEAVE-ONE-OUT CROSS VALIDATION


#MSE.LOOCV <- vector()
#for(j in 40:50) {
#  SE <- vector()
#for(i in 1:42) {
#  train <- combined_data[-i,c(1:8)]
#  test <- combined_data[i,c(1:8)]
  
  
  ####################
#  vcm.fit_cv <- vcm(y = train$MonthlyCOVIDRate, X = train[,c(3:8)],
 #              grid = train$Month, t = train$Month, bandwidth = j)
  
  ##### pred ###############
  
  #num.grid <- length(train$Month)
  #n <- nrow(train[,c(3:8)])
  #k <- 1 + ncol(train[,c(3:8)])
  #y.vec <- matrix(train$MonthlyCOVIDRate, length(train$MonthlyCOVIDRate), 1)
  #X.mat <- as.matrix(cbind("intercept" = 1, train[,c(3:8)]))
  #coef.res <- se.res <- matrix(NA, num.grid, k)
  #mse <- 0
  #for(p in 1:num.grid){
    #U.star <- as.matrix(diag(train$Month - train$Month[p], n, n))
    #Gamma.mat <- cbind(X.mat, U.star %*% X.mat)
    #wts <- my.w((train$Month - train$Month[p])/j, type)/j
    #W <- as.matrix(diag(wts, n, n))
    #est.beta <- solve(t(Gamma.mat) %*% W %*% Gamma.mat) %*%
    #  (t(Gamma.mat) %*% W %*% y.vec)
   # est.beta <- diag(1, k, 2 * k) %*% est.beta
  #  coef.res[p, ] <- est.beta
  
 # }
  #vcm.pred_cv <- colMeans(vcm.fit_cv$coef) %*% t(as.matrix(cbind("intercept" = 1, test[,c(3:8)])))
  #############
  #lm.fit <- lm(Sales ~ TV + Radio + Newspaper, data=train)
  #lm.pred <- predict(lm.fit, newdata=test)
  #as.matrix(cbind("intercept" = 1, X))
  #colMeans(vcm.fit$coef)
  
#  SE.i <- (vcm.pred_cv - test$MonthlyCOVIDRate)^2
#  SE <- rbind(SE, SE.i)
#}
# nam <- paste("MSE.LOOCV.", j, sep = "")
# assign(nam, mean(SE))

#MSE.LOOCV.j <- mean(SE)
#MSE.LOOCV <- rbind(MSE.LOOCV, MSE.LOOCV.j)
#}

```




```{r}
# Fitting a VCM
vcm.fit <- vcm(y = combined_data$MonthlyCOVIDRate, X = combined_data[, c(3:8)],
               grid = combined_data$Month, t = combined_data$Month, bandwidth = 40)

# Estimated coefficients and the corresponding 95% confidence intervals
upper <- vcm.fit$coef + 1.96 * sqrt(vcm.fit$se)
lower <- vcm.fit$coef - 1.96 * sqrt(vcm.fit$se)

# Averaged results
colMeans(vcm.fit$coef)
colMeans(lower)
colMeans(upper)

```

```{r}
#vcm.fit # mean square error, coefficients for each of the 42 months for intercept and then 4 seasonal terms and flu and rsv monthly rates,  standard error for each of the coefficients
```


```{r}
vcm_bb <- data.frame(bandwidth = c(30:800), mse = rep(NA, 771) )
for (j in 30:800) {
vcm.fit <- vcm(y = combined_data$MonthlyCOVIDRate, X = combined_data[, c(3:8)],
               grid = combined_data$Month, t = combined_data$Month, bandwidth = j)
vcm_bb[j-29,2] <- vcm.fit$mse
}
```


```{r}
plot(x = vcm_bb$bandwidth, y = vcm_bb$mse, ylim = c(563,570))
```


```{r}
# Fitting a VCM
vcm.fit <- vcm(y = combined_data$MonthlyCOVIDRate, X = combined_data[, c(3:8)],
               grid = combined_data$Month, t = combined_data$Month, bandwidth = 800)

# Estimated coefficients and the corresponding 95% confidence intervals
upper <- vcm.fit$coef + 1.96 * sqrt(vcm.fit$se)
lower <- vcm.fit$coef - 1.96 * sqrt(vcm.fit$se)

# Averaged results
colMeans(vcm.fit$coef)
colMeans(lower)
colMeans(upper)
vcm.fit$mse
```


```{r}
# Plotting
par(mfrow = c(2, 4))
plot(combined_data$Month, vcm.fit$coef[, 1], type = "l", ylim = c(- 80, 200),
     xlab = "month", ylab = "coefficient", main = "(intercept)")
lines(combined_data$Month, upper[, 1], lty = 2)
lines(combined_data$Month, lower[, 1], lty = 2)
for(j in 2:5){
  plot(combined_data$Month, vcm.fit$coef[, j], type = "l", ylim = c(- 200, 200),
       xlab = "month", ylab = "coefficient", main = paste("s", j - 1, sep = ""))
  lines(combined_data$Month, upper[, j], lty = 2)
  lines(combined_data$Month, lower[, j], lty = 2)
}
plot(combined_data$Month, vcm.fit$coef[, 6], type = "l", ylim = c(- 10, 10),
     xlab = "month", ylab = "coefficient", main = "x (flu rates)")
lines(combined_data$Month, upper[, 6], lty = 2)
lines(combined_data$Month, lower[, 6], lty = 2)

plot(combined_data$Month, vcm.fit$coef[, 7], type = "l", ylim = c(- 10, 10),
     xlab = "month", ylab = "coefficient", main = "x (rsv rates)")
lines(combined_data$Month, upper[, 7], lty = 2)
lines(combined_data$Month, lower[, 7], lty = 2)

```

```{r}
#combined_data <- combined_data[,-10]
#combined_data$predicted_mon_cov_rate <- 0
```


```{r}
# mse
predictor.mat <- as.matrix(cbind("intercept" = 1, combined_data[,3:8]))
predicted_val <- matrix(NA, 60, 1)
#for(i in 1:nrow(combined_data)) {
#predicted_val[i,] <- predictor.mat[i,] %*% vcm.fit$coef[i,]
  
#}
#mean((combined_data$MonthlyCOVIDRate - predicted_val)^2)
```


```{r}

```



```{r}
#sum( predictor.mat[10,] * vcm.fit$coef[10,])
#predictor.mat[10,] %*% vcm.fit$coef[10,]
```



```{r}

# ESTIMATION: VCM with package
library(mgcv)
gam_fit <- gam(MonthlyCOVIDRate ~ s(Month, k = 14) + # time-varying intercept
             s(Month, by = s1, k = 14) +
             s(Month, by = s2, k = 14) +
             s(Month, by = s3, k = 14) +
             s(Month, by = s4, k = 14) +
             s(Month, by = MonthlyFluRate, k = 14) +
           s(Month, by = MonthlyRSVRate, k = 14),
           data = combined_data, family = gaussian)
coef(gam_fit)
plot(gam_fit, pages = 1)

#gam_fit$coefficients


```

```{r}
par(mfrow = c(2, 4))
plot(gam_fit, select = 1,     xlab = "month", ylab = "coefficient", main = "(intercept)")
plot(gam_fit, select = 2,     xlab = "month", ylab = "coefficient", main = "s1")
plot(gam_fit, select = 3,     xlab = "month", ylab = "coefficient", main = "s2")
plot(gam_fit, select = 4,     xlab = "month", ylab = "coefficient", main = "s3")
plot(gam_fit, select = 5,     xlab = "month", ylab = "coefficient", main = "s4")
plot(gam_fit, select = 6, ylim = c(-10, 10),     xlab = "month", ylab = "coefficient", main = "x1 (flu rates)")
plot(gam_fit, select = 7, ylim = c(-10, 10),     xlab = "month", ylab = "coefficient", main = "x2 (RSV rates)")

```



```{r}
#gam.check(gam_fit)
gam.check(gam(MonthlyCOVIDRate ~ s(Month, k = 14) + # time-varying intercept
             s(Month, by = s1, k = 14) +
             s(Month, by = s2, k = 14) +
             s(Month, by = s3, k = 14) +
             s(Month, by = s4, k = 14) +
             s(Month, by = MonthlyFluRate, k = 14) +
           s(Month, by = MonthlyRSVRate, k = 14),
           data = combined_data, family = gaussian))


```

```{r}
k_values <- c(4, 5, 13, 14)
results_k <- data.frame()

for (k in k_values) {
  model <- gam(MonthlyCOVIDRate ~ s(Month, k = k) + # time-varying intercept
             s(Month, by = s1, k = k) +
             s(Month, by = s2, k = k) +
             s(Month, by = s3, k = k) +
             s(Month, by = s4, k = k) +
             s(Month, by = MonthlyFluRate, k = k) +
           s(Month, by = MonthlyRSVRate, k = k),
           data = combined_data, family = gaussian)
  aic <- AIC(model)
  results_k <- rbind(results_k, data.frame(k = k, AIC = aic))
}

print(results_k)
```

With k there are a lot of similar values near each other 1 through 4 close to 5, then there are peaks and dropoffs.


```{r}
#results_gam <- data.frame()
#for (k in 1:2) {

#loo_errors_gam <- numeric(nrow(combined_data))
#for (i in 1:nrow(combined_data)) {
  
  # 1. Split the data into training and test sets by leaving out the i-th observation
#  data_train <- combined_data[-i, ]   # Training data (leave out the i-th row)
#  data_test <- combined_data[i, ]  # Test data (just the i-th row)
  
  # 2. Fit a varying coefficient model (GAM) on the training data
  # 's(X, by = t)' means the coefficient of X varies with respect to t
 # model_train <- gam(Y ~ s(X, by = t) + s(t), data = data_train)
#  model_train <- gam(MonthlyCOVIDRate ~ s(Month, k = k) + # time-varying intercept
#                   s(Month, by = s1, k = k) +
#                   s(Month, by = s2, k = k) +
#                   s(Month, by = s3, k = k) +
#                   s(Month, by = s4, k = k) +
#                   s(Month, by = MonthlyFluRate, k = k) +
#                   s(Month, by = MonthlyRSVRate, k = k),
#                 data = data_train, family = gaussian)
  # 3. Make predictions for the left-out observation (the i-th one)
#  pred <- predict(model_train, newdata = data_test)
  
  # 4. Calculate the error for this fold (e.g., Mean Squared Error)
#  loo_errors_gam[i] <- (pred - data_test$MonthlyCOVIDRate)^2  # Squared error for LOO
  
#}

# 5. Compute overall error (Mean Squared Error)
#GAM_mean_loo_error <- mean(loo_errors_gam)
#results_gam <- rbind(results_gam, data.frame(k = k, gam_mse = GAM_mean_loo_error))
#}
#print(paste("GAM Mean LOO Error (MSE):", GAM_mean_loo_error))
```

```{r}
results_gam2 <- data.frame()
for (k in 8:9) {

loo_errors_gam <- numeric(nrow(combined_data))
for (i in 1:nrow(combined_data)) {
  
  # 1. Split the data into training and test sets by leaving out the i-th observation
  data_train <- combined_data[-i, ]   # Training data (leave out the i-th row)
  data_test <- combined_data[i, ]  # Test data (just the i-th row)
  
  # 2. Fit a varying coefficient model (GAM) on the training data
  # 's(X, by = t)' means the coefficient of X varies with respect to t
 # model_train <- gam(Y ~ s(X, by = t) + s(t), data = data_train)
  model_train <- gam(MonthlyCOVIDRate ~ s(Month, k = k) + # time-varying intercept
                   s(Month, by = s1, k = k) +
                   s(Month, by = s2, k = k) +
                   s(Month, by = s3, k = k) +
                   s(Month, by = s4, k = k) +
                   s(Month, by = MonthlyFluRate, k = k) +
                   s(Month, by = MonthlyRSVRate, k = k),
                 data = data_train, family = gaussian)
  # 3. Make predictions for the left-out observation (the i-th one)
  pred <- predict(model_train, newdata = data_test)
  
  # 4. Calculate the error for this fold (e.g., Mean Squared Error)
  loo_errors_gam[i] <- (pred - data_test$MonthlyCOVIDRate)^2  # Squared error for LOO
  
}

# 5. Compute overall error (Mean Squared Error)
GAM_mean_loo_error <- mean(loo_errors_gam)
results_gam2 <- rbind(results_gam2, data.frame(k = k, gam_mse = GAM_mean_loo_error))
}
```




```{r}

k = 14

#gam loocv
loo_errors_gam <- numeric(nrow(combined_data))
for (i in 1:nrow(combined_data)) {
  
  # 1. Split the data into training and test sets by leaving out the i-th observation
  data_train <- combined_data[-i, ]   # Training data (leave out the i-th row)
  data_test <- combined_data[i, ]  # Test data (just the i-th row)
  
  # 2. Fit a varying coefficient model (GAM) on the training data
  # 's(X, by = t)' means the coefficient of X varies with respect to t
 # model_train <- gam(Y ~ s(X, by = t) + s(t), data = data_train)
  model_train <- gam(MonthlyCOVIDRate ~ s(Month, k = k) + # time-varying intercept
                   s(Month, by = s1, k = k) +
                   s(Month, by = s2, k = k) +
                   s(Month, by = s3, k = k) +
                   s(Month, by = s4, k = k) +
                   s(Month, by = MonthlyFluRate, k = k) +
                   s(Month, by = MonthlyRSVRate, k = k),
                 data = data_train, family = gaussian)
  # 3. Make predictions for the left-out observation (the i-th one)
  pred <- predict(model_train, newdata = data_test)
  
  # 4. Calculate the error for this fold (e.g., Mean Squared Error)
  loo_errors_gam[i] <- (pred - data_test$MonthlyCOVIDRate)^2  # Squared error for LOO
  
}

# 5. Compute overall error (Mean Squared Error)
GAM_mean_loo_error <- mean(loo_errors_gam)
print(paste("GAM Mean LOO Error (MSE):", GAM_mean_loo_error))
results_gam <- rbind(results_gam, data.frame(k = k, gam_mse = GAM_mean_loo_error) )
```
```{r}
k.check(gam_fit)
```



```{r}
# ESTIMATION: LM
lm.fit <- lm(MonthlyCOVIDRate ~ ., data = combined_data[, c(2:8)])
summary(lm.fit)$coef[, 1]
summary(lm.fit)$coef[, 1] - 1.96 * summary(lm.fit)$coef[, 2]
summary(lm.fit)$coef[, 1] + 1.96 * summary(lm.fit)$coef[, 2]


```


```{r}
lm.fit$coefficients
lm.fit$residuals
lm.fit$fitted.values
```

```{r}
loo_errors_lm <- numeric(nrow(combined_data))
for (i in 1:nrow(combined_data)) {
  
  # 1. Split the data into training and test sets by leaving out the i-th observation
  data_train <- combined_data[-i, ]   # Training data (leave out the i-th row)
  data_test <- combined_data[i, ]  # Test data (just the i-th row)
  
  # 2. Fit a varying coefficient model (GAM) on the training data
  # 's(X, by = t)' means the coefficient of X varies with respect to t
  # model_train <- gam(Y ~ s(X, by = t) + s(t), data = data_train)
  
  model_train <- lm(MonthlyCOVIDRate ~ ., data = data_train[, c(2:8)])
  
  
  # 3. Make predictions for the left-out observation (the i-th one)
  pred <- predict(model_train, newdata = data_test[, c(2:8)])
  
  # 4. Calculate the error for this fold (e.g., Mean Squared Error)
  loo_errors_lm[i] <- (pred - data_test$MonthlyCOVIDRate)^2  # Squared error for LOO
  
}

# 5. Compute overall error (Mean Squared Error)
lm_mean_loo_error <- mean(loo_errors_lm)
print(paste("LM Mean LOO Error (MSE):", lm_mean_loo_error))
```


```{r}
summary(lm.fit)
```


```{r}
#tvreg lm

tvReg::bw(x = combined_data[,c(3:8)], y = combined_data$MonthlyCOVIDRate, z = combined_data$Month,
         est = 'll',  tkernel = "Epa")

tvlm <- tvReg::tvLM( MonthlyCOVIDRate ~ s1 + s2 + s3 + s4 + MonthlyFluRate + MonthlyRSVRate, z = combined_data$Month,
             data = combined_data, bw = NULL, cv.block = 0, est = 'll', tkernel = 'Epa')

tvlm$coefficients
tvlm$residuals
tvlm$bw
```

```{r}
mean(tvlm$residuals^2)
```



```{r}
loo_errors_tvlm <- numeric(nrow(combined_data))
for (i in 1:nrow(combined_data)) {
  
  # 1. Split the data into training and test sets by leaving out the i-th observation
  data_train <- combined_data[-i, ]   # Training data (leave out the i-th row)
  data_test <- combined_data[i, ]  # Test data (just the i-th row)
  
  # 2. Fit a varying coefficient model (GAM) on the training data
  # 's(X, by = t)' means the coefficient of X varies with respect to t
  # model_train <- gam(Y ~ s(X, by = t) + s(t), data = data_train)
  model_train <- tvReg::tvLM( MonthlyCOVIDRate ~ s1 + s2 + s3 + s4 + MonthlyFluRate + MonthlyRSVRate, z = data_train$Month,
                              data = data_train,  bw = 56, est = 'll', tkernel = 'Epa')
  # 3. Make predictions for the left-out observation (the i-th one)
  pred <- predict(model_train, newdata = as.matrix(data_test[,3:8]), newz = data_test$Month )
  
  # 4. Calculate the error for this fold (e.g., Mean Squared Error)
  loo_errors_tvlm[i] <- (pred - data_test$MonthlyCOVIDRate)^2  # Squared error for LOO
  
}

# 5. Compute overall error (Mean Squared Error)
tvlm_mean_loo_error <- mean(loo_errors_tvlm)
print(paste("TVLM Mean LOO Error (MSE):", tvlm_mean_loo_error))

```
setting bandwidth to constant 56 instead of recalculating bandwidth each iteration
resulted in a lower mse from 257 to 250

```{r}
# Estimation: Random Forest

library(randomForest)
RF.fit <- randomForest(MonthlyCOVIDRate ~ ., combined_data[,c(2:8)])

```

```{r}
varImpPlot(RF.fit)
```

```{r}
library(caret)
library(randomForest)

set.seed(123)

tune_grid <- expand.grid(
  .mtry = 2:5
)

control <- trainControl(method = "LOOCV")

rf_caret <- train(
  MonthlyCOVIDRate ~ .,
  data = combined_data[,2:8],
  method = "rf",
  tuneGrid = tune_grid,
  trControl = control,
  ntree = 300
)

print(rf_caret)
plot(rf_caret)
```

```{r}
RF.fit <- randomForest(MonthlyCOVIDRate ~ ., combined_data[,c(2:8)], mtry = 2, ntree = 300 )
```

```{r}
varImpPlot(RF.fit)
```


```{r}
set.seed(123)
loo_errors_rf <- numeric(nrow(combined_data))
for (i in 1:nrow(combined_data)) {
  
  # 1. Split the data into training and test sets by leaving out the i-th observation
  data_train <- combined_data[-i, ]   # Training data (leave out the i-th row)
  data_test <- combined_data[i, ]  # Test data (just the i-th row)
  
  # 2. Fit a varying coefficient model (GAM) on the training data
  # 's(X, by = t)' means the coefficient of X varies with respect to t
  # model_train <- gam(Y ~ s(X, by = t) + s(t), data = data_train)
  
  model_train <- randomForest(MonthlyCOVIDRate ~ ., data = data_train[, c(2:8)] )
  
  
  # 3. Make predictions for the left-out observation (the i-th one)
  pred <- predict(model_train, newdata = data_test[, c(2:8)])
  
  # 4. Calculate the error for this fold (e.g., Mean Squared Error)
  loo_errors_rf[i] <- (pred - data_test$MonthlyCOVIDRate)^2  # Squared error for LOO
  
}

# 5. Compute overall error (Mean Squared Error)
rf_mean_loo_error <- mean(loo_errors_rf)
print(paste("RF Mean LOO Error (MSE):", rf_mean_loo_error))
```
random forest as we will see below does well predict data that it was trained on, but when the data is new the random forest is having a relatively harder time compared to tvreg and gam

parameter tuning made the mse go up.

```{r}
mse_loocv_results <- data.frame(Method = c('LM', 'RF', 'VCM', 'TVLM'),
                                LOOCV_MSE = c(384.5461,475.3656,149.7082,250.7947),
                                MSE = c(296.1591, 108.6397, 25.8242, 153.7098))
mse_loocv_results <- mse_loocv_results[order(-mse_loocv_results$LOOCV_MSE),]
```


```{r, message = FALSE, warning = FALSE}
set.seed(123)
# COMPARISON: in terms of predictive abilities
res.pred <- matrix(NA, length(combined_data$Month), 4)
lm.fit <- lm(MonthlyCOVIDRate ~ ., data = combined_data[, c(2:8)])
RF.fit <- randomForest(MonthlyCOVIDRate ~ ., combined_data[,c(2:8)])
res.pred[, 1] <- predict(lm.fit, newdata = combined_data[, c(2:8)])
res.pred[, 2] <- predict(RF.fit, newdata = combined_data[,c(2:8)]) 
  b <- gam(MonthlyCOVIDRate ~ s(Month, k = 14) + # time-varying intercept
             s(Month, by = s1, k = 14) +
             s(Month, by = s2, k = 14) +
             s(Month, by = s3, k = 14) +
             s(Month, by = s4, k = 14) +
             s(Month, by = MonthlyFluRate, k = 14) +
             s(Month, by = MonthlyRSVRate, k = 14),
           data = combined_data, family = gaussian)
  res.pred[, 3] <- predict(b, newdata = combined_data[, c(1:8)])

  tvlm <- tvReg::tvLM( MonthlyCOVIDRate ~ s1 + s2 + s3 + s4 + MonthlyFluRate + MonthlyRSVRate, z = combined_data$Month,
             data = combined_data, bw = NULL, cv.block = 0, est = 'll', tkernel = 'Epa')
  
  res.pred[,4] <- predict(tvlm, newdata = as.matrix(combined_data[,3:8]), newz = combined_data$Month)
  
# Figures 4 and 5
library(dplyr)
library(tidyr)

n <- nrow(combined_data)

# (A) Data for the time-series plot (Figure 5)
df_time <- data.frame(
  Month  = combined_data$Month,
  Actual = combined_data$MonthlyCOVIDRate,
  LM     = res.pred[, 1],
  RF     = res.pred[, 2],
  VCM    = res.pred[, 3],
  TVLM   = res.pred[, 4]
)

# Convert to long format for plotting lines over time
df_time_long <- pivot_longer(df_time, cols = c("Actual", "LM", "RF", "VCM", "TVLM"),
                             names_to = "Model", values_to = "Value")
                             
# (B) Data for the scatter plot (Figure 4)
# Here, we want to compare observed vs. fitted values for LM and VCM.
df_scatter <- data.frame(
  Observed = rep(combined_data$MonthlyCOVIDRate, times = 4),
  Fitted   = c(res.pred[, 1], res.pred[, 2], res.pred[, 3], res.pred[, 4]),
  Model    = rep(c("LM", "RF", "VCM", "TVLM"), each = n)
)

# Figure 4: Scatter Plots of Observed vs. Fitted
library(ggplot2)
ggplot(df_scatter, aes(x = Observed, y = Fitted)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  facet_wrap(~ Model, nrow = 1) +
  xlim(5, 130) +
  ylim(-10, 100) +
  labs(x = "Observed", y = "Fitted", title = "Observed vs. Fitted") +
  theme_minimal()

# Figure 5: Time Series of Actual, LM, and VCM Values
ggplot() +
  # Layer for "Actual" (thicker line)
  geom_line(data = filter(df_time_long, Model == "Actual"),
            aes(x = Month, y = Value, color = Model, linetype = Model),
            size = 2) +
  # Layer for "LM" and "VCM" (thinner lines)
  geom_line(data = filter(df_time_long, Model != "Actual"),
            aes(x = Month, y = Value, color = Model, linetype = Model),
            size = 1) +
  scale_color_manual(values = c("Actual" = "black", "LM" = "red", "RF" = "green3", "VCM" = "blue", "TVLM" = "pink")) +
  scale_linetype_manual(values = c("Actual" = 1, "LM" = 2, "RF" = 3, "VCM" = 4, "TVLM" = 5)) +
  ylim(5, 130) +
  labs(x = "Month", y = "COVID Rates", title = "COVID Rates over Time") +
  theme_minimal() +
  theme(legend.position = c(0.85, 0.85))

# Table 2
c(mean((res.pred[, 1] - combined_data$MonthlyCOVIDRate)^2), mean((res.pred[, 2] - combined_data$MonthlyCOVIDRate)^2),
  mean((res.pred[, 3] - combined_data$MonthlyCOVIDRate)^2), mean((res.pred[, 4] - combined_data$MonthlyCOVIDRate)^2))

```



```{r}
#combined_data$Month[-4]
```

```{r}
tvlm$bw
# bandwidth still 56
```


```{r}
# LEAVE-ONE-OUT CROSS VALIDATION


#MSE.LOOCV <- vector()
#for(j in 40:600) {
#  SE <- vector()
#for(i in 1:42) {
  #train <- combined_data[-i,c(1:8)]
  #test <- combined_data[i,c(1:8)]
  
  
  ####################
  #vcm.fit_cv <- vcm(y = train$MonthlyCOVIDRate, X = train[,c(3:8)],
  #             grid = train$Month, t = train$Month, bandwidth = j)
 # vcm.pred_cv <- colMeans(vcm.fit_cv$coef) %*% t(as.matrix(cbind("intercept" = 1, test[,c(3:8)])))
  #############
  #lm.fit <- lm(Sales ~ TV + Radio + Newspaper, data=train)
  #lm.pred <- predict(lm.fit, newdata=test)
  #as.matrix(cbind("intercept" = 1, X))
  #colMeans(vcm.fit$coef)
  
#  SE.i <- (vcm.pred_cv - test$MonthlyCOVIDRate)^2
#  SE <- rbind(SE, SE.i)
#}
# nam <- paste("MSE.LOOCV.", j, sep = "")
# assign(nam, mean(SE))

#MSE.LOOCV.j <- mean(SE)
#MSE.LOOCV <- rbind(MSE.LOOCV, MSE.LOOCV.j)
#}

```



```{r}
#SE <- vector()
#for(i in 1:42) {
  #train <- combined_data[-i,c(1:8)]
  #test <- combined_data[i,c(1:8)]
  
  
  ####################
  #vcm.fit_cv <- vcm(y = train$MonthlyCOVIDRate, X = train[,c(3:8)],
  #             grid = train$Month, t = train$Month, bandwidth = 50)
 # vcm.pred_cv <- colMeans(vcm.fit_cv$coef) %*% t(as.matrix(cbind("intercept" = 1, test[,c(3:8)])))
  #############
  #lm.fit <- lm(Sales ~ TV + Radio + Newspaper, data=train)
  #lm.pred <- predict(lm.fit, newdata=test)
  #as.matrix(cbind("intercept" = 1, X))
  #colMeans(vcm.fit$coef)
  
#  SE.i <- (vcm.pred_cv - test$MonthlyCOVIDRate)^2
#  SE <- rbind(SE, SE.i)
#}
# nam <- paste("MSE.LOOCV.", j, sep = "")
# assign(nam, mean(SE))

#MSE.LOOCVz <- mean(SE)
#MSE.LOOCVz
```

```{r}
#plot(x=c(40:600), y = MSE.LOOCV)
#abline(h = 452, col = 'red')
```
```{r}
#msecvloo <- data.frame(bandwidth = c(40:600), mse = MSE.LOOCV)
#ggplot(msecvloo, aes(bandwidth,mse)) +
#  geom_point() +
#  scale_x_continuous(breaks = pretty(msecvloo$bandwidth, n=10)) +
#  scale_y_continuous(breaks = pretty(msecvloo$mse, n=10))+
#  geom_abline(intercept = 452, slope = 0, col = 'red')

```
```{r}
#ggplot(msecvloo, aes(bandwidth,mse)) +
#  geom_point() +
#  scale_x_continuous(breaks = pretty(msecvloo$bandwidth, n=10)) +
#  scale_y_continuous(breaks = pretty(msecvloo$mse, n=10))+
#  geom_abline(intercept = 452, slope = 0, col = 'red')+
#  ylim(451,455)
```

